---
title: "1. Getting started with coconatfly"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Getting started with coconatfly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(fafbseg.use_static_celltypes = TRUE)
```

## Setup for brain connectome data.

At the time of writing the main datasets of interest for coconatfly will be the hemibrain and flywire. Installing coconatfly will automatically install the natverse packages required for analysis. However, a few extra steps are still required.

```{r install-brain, eval=FALSE}
# first install coconatfly
if(!requireNamespace('natmanager'))
  install.packages('natmanager')
natmanager::install(pkgs = 'coconatfly')
# then get flywire data
fafbseg::download_flywire_release_data()
```

If you have not already done so, you will also need to record a neuprint token to access hemibrain data. Go to <https://neuprint.janelia.org/account>, copy your Auth Token and then edit your R environ file to do

```{r setting token, eval=FALSE}
usethis::edit_r_environ()
```

and add a line like this:

```         
neuprint_token="eyJhbH0lc9OqzJ6evvcJ23KmIi.eyJlb6B8nUg8senJO1ZVzHb6uQ1fvncAiWdFcTIwTQb0xOcxa3TjKHVdLZ5OAtHzjWJJLqpQjiTikWryIfgUHCKoq31RSdDMqD7mah0RWCFaEPrTDdPm7OfRBizNdoqyyWAE61g9FWb2CAh8kAfcsb3ylVG3MwFOYpX91ayWxtp1prfjtm9dte9mFddGdWdr8444Rrun8cLksyELl93utdnGAQReSgiNy-jDLujyRCSRCAOq23X049gRA9QtJcOKCDtdiFyQhMKzLk8EISt7Cnn9AS1"
```

Replacing with your auth token and being sure that is all on one line with a line break after the last quotation mark.

## Set up for VNC data

For VNC data, the main datasets are FANC and MANC. MANC data will be available if you have set up your neuprint access token as described in the brain section.
For FANC data you will need a separate *CAVE* access token. Essentially this is covered by:

```{r install-vnc, eval=FALSE}
fancr::fanc_set_token()
# check all is good
fancr::dr_fanc()
```

## Dataset status

You can get a useful overview of the current status of your access to different
datasets with the `dr_coconatfly()` command.

```{r, message=FALSE}
library(coconatfly)
dr_coconatfly()
```


## Brain example

Ok with that in place, let's check if we can use connectome data, starting
with the brain.

```{r setup, message=FALSE}
library(coconatfly)
library(dplyr)
```

We use `cf_ids()` to specify neurons in different datasets:
```{r}
cf_ids(hemibrain = 'ORN_DA2', expand = TRUE)
```

We can ask for the same specific cell type across datasets, here
MBON01, mushroom body output neuron 1 (aka M6 or MBON-gamma5beta'2a).

```{r}
cf_meta(cf_ids('MBON01', datasets = c("flywire", "hemibrain")))
```
We can do a regular expression query returning all MBONs:
```{r}
mbonmeta=cf_meta(cf_ids('/type:MBON.+', datasets = c("flywire", "hemibrain")))
mbonmeta
```
and then do some simple summaries on the metadata returned by `cf_meta()`:

```{r}
mbonmeta %>% 
  mutate(dss=paste0(abbreviate_datasets(dataset), side)) %>% 
  with(table(type, dss))
```
In general things look quite consistent across sides and datasets (although some neurons are of course missing from the LHS of the hemibrain).

### Connectivity clustering

We can then then inspect the situation for MBON25 and MBON34 which have 
been combined into a single compound type in FlyWire but were regarded as two
distinct types in hemibrain. Notice that we pipe ` %>% ` the dataframe containing metadata to the cf_cosine_plot function via a `filter` expression which picks out our cell type of interest as well as few
similarly numbered ones. We use output connectivity clustering since we suspect the downstream cell type interactions are more diagnostic for 

```{r}
mbonmeta %>% 
  filter(grepl("MBON(25|34|24|26|33|35)", type)) %>% 
  cf_cosine_plot(partners = 'out')
```

