---
title: "2. Metadata columns and harmonisation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Metadata columns and harmonisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

When querying metadata across multiple connectome datasets with `cf_meta()`,
*coconatfly* returns a standardised set of columns. This vignette documents these
columns and explains how values are harmonised across datasets.

```{r setup, message=FALSE}
library(coconatfly)
library(dplyr)
```

## Standard columns

The `cf_meta()` function returns these standard columns:

| Column | Description |
|--------|-------------|
| `id` | Neuron identifier (dataset-specific) |
| `key` | Unique identifier across all datasets (format: `dataset:id`) |
| `dataset` | Source dataset name |
| `class` | Top-level cell class (optionally harmonised) |
| `subclass` | Cell subclass |
| `type` | Cell type name, the finest classification level |
| `instance` | Individual neuron properties (typically formatted as `{type}_{side}`) |
| `side` | Normalised to L/R/M (left/right/midline) or NA |
| `sex` | M or F, based on the dataset |
| `tissue` | brain, vnc, or cns depending on dataset |
| `group` | Numeric grouping of related neurons within dataset |

### Example: MBON01 across datasets

```{r mbon01}
mbon01 <- cf_meta(cf_ids("MBON01", datasets = c("flywire", "malecns")))
mbon01 %>%
  select(id, class, type, side, sex, tissue, dataset, key)
```

## Dataset properties

### Sex

The `sex` column reflects the sex of the animal from which the dataset was
derived:

| Dataset | Sex | Description |
|---------|-----|-------------|
| flywire | F | Female adult brain |
| hemibrain | F | Female adult brain |
| opticlobe | M | Male optic lobe |
| malecns | M | Male central nervous system |
| manc | M | Male adult nerve cord |
| fanc | F | Female adult nerve cord |
| banc | F | Female adult nerve cord |
| yakubavnc | M | Male D. yakuba VNC |

### Tissue

The `tissue` column indicates the region of the nervous system:

- `brain`: Brain datasets (flywire, hemibrain, opticlobe)
- `vnc`: Ventral nerve cord datasets (manc, fanc, yakubavnc)
- `cns`: Complete CNS datasets (malecns, banc)

## Side normalisation

The `side` column is normalised to single uppercase letters:

- `L`: Left
- `R`: Right
- `M`: Midline (or center)
- `NA`: Unknown or not applicable

Input values like "left", "Left", "L", "l" are all normalised to "L".

## Superclass/class field names

Most datasets have a multi-level annotation hierarchy from very broad terms to 
terminal cell types, which represent the finest level of conservation across
datasets. Different datasets use different conventions for names and values of 
these fields. We have already noted that `type` is our default name for the 
finest level classification. The flywire data model calls this `cell_type` but
we always remap the column name to type.

For malecns, the top level in the `class` hierarchy is `superclass`, while
flywire uses `super_class` and manc uses `class`. For historical reasons,
the default is to rename all of these to `class`.
You can opt to have the top level column as `superclass`. The renaming can be
controlled by the `coconatfly.use_superclass` package option. 

```{r superclass, eval=FALSE}
options(coconatfly.use_superclass = TRUE)
# or
cf_meta(ids, use_superclass = TRUE)
```

This currently
defaults to FALSE, but we expect a future version to use this by default, more
closely emulating the male cns.

- `class` -> `superclass` (values like ascending_neuron)
- `subclass` -> `class` (values like ALPN)
- `subsubclass` -> `subclass`

## Class harmonisation

Even when the columns have the same name, values may differ. For example, 
ascending neurons that project from the VNC to the brain might be labelled:

- malecns: `ascending_neuron`
- manc: `ascending neuron`
- flywire: `ascending`

in different datasets.

### Enabling harmonisation

By default, class values are **not** harmonised, preserving the original
values in each dataset. To enable harmonisation to malecns-style values,
use the `harmonise_class` parameter or set a global option:

```{r harmonise-param-options, eval=FALSE}
cf_meta(ids, harmonise_class = TRUE)

options(coconatfly.harmonise_class = TRUE)
cf_meta(ids)
```

A future version of coconatfly will make harmonisation the default.

### Example: Ascending neurons

Without harmonisation, class values differ between datasets:
```{r an-no-harmonise}
an_meta <- cf_meta(cf_ids("AN07B004", datasets = c("manc", "malecns")))
an_meta %>%
  select(type, class, side, dataset)
```

With harmonisation enabled:
```{r an-harmonise}
an_meta_h <- cf_meta(cf_ids("AN07B004", datasets = c("manc", "malecns")),
                     harmonise_class = TRUE)
an_meta_h %>%
  select(type, class, side, dataset)
```

### Harmonisation rules

When `harmonise_class = TRUE`, class values are mapped to malecns conventions:

| Pattern | Harmonised value |
|---------|------------------|
| ascending (+ neuron) | ascending_neuron |
| descending (+ neuron) | descending_neuron |
| sensory + ascending | sensory_ascending |
| sensory + descending | sensory_descending |
| efferent + ascending | efferent_ascending |
| sensory | {domain}_sensory |
| motor | {domain}_motor |
| intrinsic, central | {domain}_intrinsic |
| efferent | {domain}_efferent |
| optic | ol_intrinsic |
| visual projection | visual_projection |
| visual centrifugal | visual_centrifugal |
| endocrine | endocrine |

Where `{domain}` is `cb` (central brain), `vnc`, or `ol` (optic lobe) based on
the dataset.

