---
title: "TuTu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TuTu}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(coconatfly)
```

TuTu neurons are a class of visual neurons that connect the two optic tubercles. From what we know of them they are likely to be responsive to small field objects e.g. other flies.

You can read more about them on VFB, where they are called [`adult tubercle-tubercle neuron` [FBbt_00051062]](https://v2.virtualflybrain.org/org.geppetto.frontend/geppetto?id=FBbt_00051062&i=VFB_00101567,VFB_jrchk7cz).

The following neurons are recorded in the hemibrain and FAFB-FlyWire. Note the use of a regular expression to match against the type field.

```{r}
tutus=cf_meta(cf_ids('/type:TuTu.*', datasets = c("hemibrain", "flywire")))
tutus
```

```{r}
tutus %>% 
  count(dataset, type, side)
```
So we can see the hemibrain TuTuB neurons have been typed into two subtypes, but the flywire neurons have not. First of all does
cosine clustering in the hem

```{r}
tutus %>% 
  cf_cosine_plot()
```

So this is interesting at a couple of levels. First we can see that the TuTuB
hemibrain neurons separate into two groups. This is true even though things are
a bit messy for the
`TuTuB_a_hbL` neuron which will almost certainly be heavily truncated as it has 
its soma on the left hand (missing) side of the hemibrain.

We can also see the flywire neurons also split up nicely. So we have two groups
of four neurons each. Each group contains one neuron from each hemisphere x dataset combination. So we can update our flywire cell typing to match the hemibrain
TuTuB_a and TuTuB_b groups.


### TuTuA

However, interestingly we can also see that the TuTuA group splits very strongly
in flywire into two groups that are not evident in the hemibrain. To me this
strongly suggests that some property of the bilateral connectivity defines this
difference. 

We can try seeing if we just use the input connectivity to cluster if things
resolve:

```{r}
tutus %>% 
  filter(grepl("TuTuA", type)) %>% 
  cf_cosine_plot(partners = 'in')
```

Hmm the situatuion is still the same. Since the two flywire groups are so well-separated
this is almost certainly a real split. However since the hemibrain does not give
any data to support this, I will not record the split for now. I strongly suspect that
we will observe this split in future connectomes though.
